Evaluating Recommender Behavior For New Users

### 2014-07-21 paper/main.tex
first pass for photo-ready. Still needs work

  As mentioned earlier, many recommender systems use a new user survey to collect the first user ratings.
  These ratings are different than normal user ratings.
  As we explain shortly, our methodology uses randomization in key places to remove this ordering effect.
  This allows us to better capture the underlying algorithm performance on ``normal'' user data.
  While this makes our methodology unsuited for applying temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.
  Therefore, we feel that our methodology is more suitable to answer our research question.

@@

  We seek to develop an understanding of how a range of different standard algorithms perform for new users.
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare UserUser\cite{resnick1994grouplens}, ItemItem\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} against two simple baseline algorithms.
  More information about these algorithms can be found in table \ref{tbl:algo}.
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

@@

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  Each algorithm was tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
  Since this evaluation was on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning from these parameters.
  Most notably, we found that the previous damping parameter used for the baselines was too large.
  We found that a value of 5 gave much better results for new user recommendation.
  Our key results are not sensitive to minor changes of algorithm parameters.
  %TODO: carefully run this wording by joe.

@@

  We use a total of eleven metrics.
  In our evaluation we found that some of these metrics are redundant, and future work should be able to use a much smaller list of metrics.
  The metrics are listed in table \ref{tbl:metrics}.
  We split our metrics into three groups roughly along our three sub research questions.

  \begin{table*}[ht!]
    \centering
    \begin{tabular}{|p{7.5em}|p{21em}|p{7em}|p{12.5em}|}
      \hline
      Metric & Description & Range & Preferred direction\\\hline
      \hline
      \multicolumn{4}{|c|}{{\bf Accuracy and Ranking Metrics}} \\\hline
      RMSE                   & Standard prediction accuracy metric \cite{handbook}. & 0 to 5 stars & Smaller is better. \\\hline
      nDCG                   & Standard ranking quality metric \cite{handbook}.     & 0 to 1       & Larger is better.  \\\hline
      \hline
      \multicolumn{4}{|c|}{{\bf Monotonic Recommendation Metrics}} \\\hline
      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher\footnotemark[1] \cite{handbook}. & 0 to 1       & Larger is better.  \\\hline
      MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}.                & 0 to 1       & Larger is better.  \\\hline
      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}.  & 0 to 1       & Smaller is better. \\\hline
      MeanRating@20          & Average rating of recommended items that are in the test set.                                       & 0 to 5 stars & Larger is better.  \\\hline
      RMSE@20                & RMSE computed only over those items that were recommended.                                          & 0 to 5 stars & Smaller is better. \\\hline
      \hline
      \multicolumn{4}{|c|}{{\bf Non-monotonic Recommender Metrics}} \\\hline
      SeenItems@20           & Count of recommended items in the test set. & 0 to 20 items & Moderate values are better. \\\hline
      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items \cite{zieglerDiversity}.& 1 to 6000 users & Moderate values are better. \\\hline
      AILS@20                & The average pairwise similarity between recommended items \cite{zieglerDiversity}.                       & -1 to 1         & Moderate values are better. \\\hline
      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.                & 0 to 12 bits    & Moderate values are better. \\\hline
    \end{tabular}
    \caption{Summary of metrics}
    \label{tbl:metrics}
  \end{table*}

@@

  \paragraph{Monotonic Recommendation Metrics}
  %TODO: kenny comment: is 20 standard size for these?
  %TODO: daniel comment: check for consistency with the user vs. users.
  These metrics evaluate the top 20 recommendations produced by the recommender.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.

  These metrics are useful because they have a well defined relation to the satisfaction of the user.
  As the metric grows larger (or smaller, depending on metric) we expect the recommendations to satisfy users more.
  This makes these metrics easy to apply.

  {\bf Precision@20}, {\bf MAP@20}, and {\bf Fallout@20} are standard information retrieval metrics.
  As we show later, we found these metrics to not be useful for this evaluation.
  Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.

  {\bf MeanRating@20} tries to establish how much users like their recommendations by measuring users average satisfaction with recommended items based on withheld ratings.
  {\bf RMSE@20} accompanies this metric and lets us know if the recommended items are likely to be incorrectly predicted, which can explain why users may or may not like their recommendations.

  \paragraph{Non-monotonic Recommendation Metrics}
  {\bf SeenItems@20}, {\bf AveragePopularity@20}, {\bf AILS@20}, and {\bf Spread@20} are all non-monotonic measures of the user's recommendations.
  These metrics are harder to apply than the monotonic measures of recommendation quality.
  Unlike the previous metrics, we expect that both of the extremes of these metrics relate to a negative user experience.

@@

  {\bf AveragePopularity@20} is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
  Broadly speaking, we expect users will prefer recommendation lists containing more novel (less popular) items.
  However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experience.

@@

  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by ziegler et al \cite{zieglerDiversity}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  We used the same item similarity metric for AILS@20 as we used in the ItemItem recommender.

  The {\bf Spread@20} metric is designed to measure how well the recommender spreads its attention across many items.
  We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
  Like diversity, however, we expect that it is not possible to reach complete spread (recommending each item an equal number of times) without making avoidably bad recommendations.
  When Spread is large we know that the recommender is recommending many items with a relatively even distribution.
  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.

@@

  \paragraph{Expectations about algorithm performance}
  % provide expectations.
  We expect that the ItemBaseline algorithm will have approximately constant performance on all metrics.
  Since the ItemBaseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better than it.
  Furthermore, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.

  The UserItemBaseline should only differ from the ItemBaseline for metrics that use the prediction value.
  For all other metrics these two provide the same responses, and we will only report results for the ItemBaseline.
  As the UserItemBaseline is still quite a simple baseline, we expect that the non-baseline algorithm will beat this baseline as well.
  That said, for small rating values it is possible that the improvements over this baseline will be very small.

  We expect that the baseline algorithms will provide the most popular recommendations, with the least diversity, and the least spread.
  We expect that any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
  Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations.

@@

  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{4.5em}|p{4.5em}|p{7.5em}|p{4.5em}|}
      \hline
      Algorithm & Prediction accuracy & Recommendation quality      & Other properties \\\hline
      ItemItem  & Poor                & Poor                        & Good             \\\hline
      UserUser  & Good                & Inconclusive                & Poor             \\\hline
      SVD       & Good                & Good                        & Good             \\\hline
    \end{tabular}
    \caption{Summary of algorithm behavior for new users}
    \label{tbl:results}
  \end{table}

@@

\section{Conclusion}

  Our work shows that a multiple metric approach to understanding recommender behavior can provide useful insights into the differences between algorithms for new users.
  Our methodology presents a list of metrics, and an evaluation framework that can be used to perform further evaluations in the future.
  We find that recommending for new users is hard and none of our algorithms can reliably beat a simple baseline for users with very few ratings.
  Once the user has a small number of ratings we find that the SVD algorithm performs best.
  This result can be useful both in designing new recommender systems, and in designing novel recommendation algorithms for new users.

>>>>

  As mentioned earlier, many recommender systems use a new user survey to collect the first user ratings.
  These ratings are different than normal user ratings.
  As we explain shortly, our methodology uses randomization in key places to remove this ordering effect.
  This allows us to better capture the underlying algorithm performance on ``normal'' user data.
  While this makes our methodology unsuited for applying temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.

Therefore, we feel that our methodology is more suitable to answer our research question.

@@

  We seek to develop an understanding of how a range of different standard algorithms perform for new users.
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare UserUser\cite{resnick1994grouplens}, ItemItem\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} against two simple baseline algorithms.

  Due to the low profile sizes involved in our evaluation, not every algorithm was able to make predictions for every user-item pair.
  In these cases, we use the UserItemBaseline prediction as a fallback.
  This ensured that our prediction based metrics were always evaluated over the same number items.
  We found this unnecessary for recommendations as each algorithm was able to make a complete recommendation with only one rating.
  Therefore our algorithms recommend only items that they can predict.
  More information about these algorithms can be found in table \ref{tbl:algo}.
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

@@

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using the Ekstrand's values as a starting point.
  We tuned based on RMSE performance.
  While tuning  we found that the previous damping parameter used for the baselines to be too large.
  We found that a value of 5 gave much better results for new user recommendation.
  Our key results are not sensitive to minor changes of algorithm parameters.

@@

  \begin{table*}[ht!]
    \centering
    \begin{tabular}{|p{7.5em}|p{21em}|p{7em}|p{12.5em}|}
      \hline
      Metric & Description & Range & Preferred direction\\\hline
      \hline
      \multicolumn{4}{|c|}{{\bf Accuracy and Ranking Metrics}} \\\hline
      RMSE                   & Standard prediction accuracy metric \cite{handbook}. & 0 to 5 stars & Smaller is better. \\\hline
      nDCG                   & Standard ranking quality metric \cite{handbook}.     & 0 to 1       & Larger is better.  \\\hline
      \hline
      \multicolumn{4}{|c|}{{\bf Monotonic Recommendation Metrics}} \\\hline
      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher\footnotemark[1] \cite{handbook}.   & 0 to 1       & Larger is better.  \\\hline
      MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}.                  & 0 to 1       & Larger is better.  \\\hline
      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{TenIsEnough}. & 0 to 1       & Smaller is better. \\\hline
      MeanRating@20          & Average rating of recommended items that are in the test set.                                         & 0 to 5 stars & Larger is better.  \\\hline
      RMSE@20                & RMSE computed only over those items that were recommended.                                            & 0 to 5 stars & Smaller is better. \\\hline
      \hline
      \multicolumn{4}{|c|}{{\bf Non-monotonic Recommender Metrics}} \\\hline
      SeenItems@20           & Count of recommended items in the test set. & 0 to 20 items & Moderate values are better. \\\hline
      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items \cite{zieglerDiversity}.& 1 to 6000 users & Moderate values are better. \\\hline
      AILS@20                & The average pairwise similarity between recommended items \cite{zieglerDiversity}.                       & -1 to 1         & Moderate values are better. \\\hline
      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.                & 0 to 12 bits    & Moderate values are better. \\\hline
    \end{tabular}
    \caption{Summary of metrics}
    \label{tbl:metrics}
  \end{table*}

  We present results from eleven metrics.
  As has been seen in previous studies \cite{herlockerEvaluating} we find that some of are metrics are redundant.
  Future work should be able to use a smaller list of metrics.
  The metrics are listed in table \ref{tbl:metrics}.
  We split our metrics into three groups roughly along our three sub research questions.

@@

  \paragraph{Monotonic Recommendation Metrics}
  %TODO: daniel comment: check for consistency with the user vs. users.
  These metrics evaluate the top 20 recommendations produced by the recommender.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.
  We refer to these as monotonic metrics as there is a clear good direction for the metric.

  {\bf Precision@20}, {\bf MAP@20}, and {\bf Fallout@20} are standard information retrieval metrics which measure the quality of the recommendations.
  As we show later, these metrics were not useful for our evaluation.
  Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.

  {\bf MeanRating@20} tries to establish how much users like their recommendations by measuring users average satisfaction with recommended items based on withheld ratings.
  {\bf RMSE@20} accompanies this metric and lets us know if the recommended items are likely to be incorrectly predicted. 
  This can help us understand why an algorithm scores well at MeanRatings@20.

  \paragraph{Non-monotonic Recommendation Metrics}
  {\bf SeenItems@20}, {\bf AveragePopularity@20}, {\bf AILS@20}, and {\bf Spread@20} are all non-monotonic measures of the user's recommendations.
  These metrics are harder to apply than the monotonic measures of recommendation quality.
  Unlike the previous metrics, we expect that users will be most satisfied by moderate values of these metrics, and that both extremes can lead to a negative user experience.

@@

  {\bf AveragePopularity@20} is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
  Broadly speaking, we expect users will prefer recommendation lists containing more novel (less popular) items.
  However, if the recommended items are too novel the user is unlikely to have heard of the items and will have no point of reference in understanding their recommendations.

@@

  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by Ziegler et al \cite{zieglerDiversity}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  We used the same item similarity metric for AILS@20 as we used in the ItemItem recommender.

  The {\bf Spread@20} metric is designed to measure how well the recommender spreads its attention across many items.
  We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
  Like diversity, however, we expect that it is not possible to reach complete spread (recommending each item an equal number of times) without making avoidably bad recommendations.
  When Spread is large we know that the recommender is recommending many items with a relatively even distribution.
  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.
  In this way spread is related to item coverage metrics.
  Spread however looks at how often each item is recommended, not just whether or not it is recommended.

@@

  \paragraph{Expectations about algorithm performance}
  % provide expectations.
  We expect that the ItemBaseline algorithm will have approximately constant performance on all metrics.
  Since the ItemBaseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better than it.
  That said, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.

  The UserItemBaseline should only differ from the ItemBaseline for metrics that use the prediction value.
  For all other metrics these two provide the same responses, and we will only report results for the ItemBaseline.
  As the UserItemBaseline is still quite a simple baseline, we expect that the non-baseline algorithm will beat this baseline as well given enough ratings.

  We expect that the baseline algorithms will provide the most popular recommendations, with the least diversity, and the least spread.
  Any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
  Furthermore, we expect that as an algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations.

@@

  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{4.5em}|p{4.5em}|p{7.5em}|p{4.5em}|}
      \hline
      {\bf Algorithm} & {\bf Prediction accuracy} & {\bf Recommendation quality} & {\bf Other properties} \\\hline
      ItemItem        & Poor                      & Poor                         & Good                   \\\hline
      UserUser        & Good                      & Inconclusive                 & Poor                   \\\hline
      SVD             & Good                      & Good                         & Good                   \\\hline
    \end{tabular}
    \caption{Summary of algorithm behavior for new users}
    \label{tbl:results}
  \end{table}

@@

\section{Conclusion}
  We find that recommending for new users is hard and none of our algorithms can reliably beat a simple baseline for users with very few ratings.
  Once the user has a small number of ratings we find that the SVD algorithm performs best.
  This result can be useful both in designing new recommender systems, and in designing novel recommendation algorithms for new users.
  We also present a methodology and list of metrics for understanding recommender behavior that can be used to extend our results to other algorithms and datasets.
